Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-01-08 sources/bmc_template_submission/bmc_article.tex
last edits

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with molecular sequence data that has never been
incorporated into any phylogenetic estimate provides the opportunity to capture
the evolutionary history of many taxa.
Here, we introduce Physcraper, a tool that automates database connections
to build on published alignments to extend phylogenetic inferences with more taxa.
Physcraper can be used to update a starting tree and single locus alignments
with public DNA data, and link the tips in these trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}.

@@

Physcraper is a Python pipeline using OpenTree's programmatic access
protocols (API's) to automatically
link phylogenies mapped to a standardized taxonomy \cite{mctavish2015phylesystem}, to alignments
from TreeBASE \cite{sanderson1994treebase}, data from GenBank \cite{benson2000genbank}.
Physcraper provides automated and reproducible workflows for updating existing phylogenetic estimates.

@@

Physcraper builds on this automated database mining concept by incorporating prior phylogenetic work and existing taxonomic domain knowledge on appropriate markers and alignment construction.
This decreases error (requiring more manual downstream processing) and eases comparison with previous phylogenetic
knowledge. 
Unlike placement approaches \cite{berger_performance_2011, matsen_pplacer_2010}, which add new taxa without modifying the input tree, Physcraper estimates all the relationships in the context of the new data.
PUMPER \cite{izquierdo2014pumper} shares these conceptual strengths, but is no longer under active development, is challenging to install and run, and has resulted in very few phylogenetic analyses since its publication.

Physcraper generates gene trees, which individually do not capture the full complexity of species' evolutionary history \cite{song2012resolving}.
However, physcraper facilitates
gathering alignments and gene trees for multiple loci from a group of interest,
that together can be used to reconstruct species trees taking into account coalescent processes with ASTRAL \cite{mirarab2014astral},
BEAST2 \cite{bouckaert2019beast}, or SVD Quartets \cite{chifman2014quartet}).
Rigorous analyses of multiple loci allows for more complex evolutionary models than analyses of large genomic data sets, and can provide better evolutionary estimates.

@@

The Physcraper workflow can be used to rapidly (in a matter of hours) to create phylogenies which can address challenges overarching both fields of ecology and evolution, such as
phylogenetically placing newly discovered species \cite{webb2010biodiversity},
curating taxonomic assignments
\cite{san2010molecular},
and generating custom trees for ecological \cite{helmus2012phylogenetic} and
evolutionary downstream analyses \cite{stoltzfus2013phylotastic}.

>>>>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with publicly available molecular sequence data that
has never been
incorporated into any phylogenetic estimate provides the opportunity to capture
the evolutionary history of many taxa.
Here, we introduce Physcraper, a tool that automates database connections
to build on published alignments to extend phylogenetic inferences with more taxa.
Physcraper can be used to update a starting tree and single locus alignments
with public DNA data, and link the tips in these trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}.

@@

Physcraper is a Python pipeline using OpenTree's programmatic access
protocols (API's) to automatically
link phylogenies mapped to a standardized taxonomy \cite{mctavish2015phylesystem}, to alignments
from TreeBASE \cite{sanderson1994treebase}, and data from GenBank \cite{benson2000genbank}.
Physcraper provides automated and reproducible workflows for updating existing phylogenetic estimates.

@@

Physcraper builds on this automated database mining concept by incorporating prior
phylogenetic work and existing taxonomic domain knowledge on appropriate markers
and alignment construction.
This decreases error (requiring less manual downstream processing) and
eases comparison with previous phylogenetic
knowledge.
Unlike phylogenetic placement approaches \cite{berger_performance_2011, matsen_pplacer_2010},
which add new taxa without modifying the input tree, Physcraper estimates all
the relationships anew in the context of the new data.
PUMPER \cite{izquierdo2014pumper} shares these conceptual strengths, but is no
longer under active development, is challenging to install and run, and has resulted
in very few phylogenetic analyses since its publication.

Physcraper generates gene trees, which individually do not capture the full complexity
of species' evolutionary history \cite{song2012resolving}.
However, Physcraper facilitates
gathering alignments and gene trees for multiple loci from a group of interest,
that together can be used to reconstruct species trees taking into account coalescent
processes with ASTRAL \cite{mirarab2014astral},
BEAST2 \cite{bouckaert2019beast}, or SVD Quartets \cite{chifman2014quartet}).
Rigorous analyses of multiple loci allows for more complex evolutionary models
than analyses of large genomic data sets, and can provide better evolutionary estimates.

@@

The Physcraper workflow can be used to rapidly (in a matter of hours) create 
phylogenies which can address challenges overarching both fields of ecology and evolution, such as
phylogenetically placing newly discovered species \cite{webb2010biodiversity},
curating taxonomic assignments
\cite{san2010molecular},
and generating custom trees for ecological \cite{helmus2012phylogenetic} and
evolutionary downstream analyses \cite{stoltzfus2013phylotastic}.
